An Effective Model of HL-60 Differentiation

### 2017-05-05 Paper_v2.tex
First draft

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained MEK-dependent activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
and the formation of an ATRA-inducible signalsome complex at the membrane which drives MAPK activation through a yet to be identified kinase activity.
While the makeup of the signalsome complex is not yet known,
we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
Signalsome formation and activity is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1, identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}, and is also involved with signalsome activity.
Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
However, this previous study developed and analyzed a complex model,
thus leaving open the critical question of what is the minimal positive feedback circuit required to drive ATRA-induced differentiation.

In this study, we explored this question using a minimal mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
The ATRA responsive signalsome-MAPK circuit was then used to drive a downstream gene expression program which encoded for the expression of functional differentiation markers.
The effective model used a novel framework which integrated logical rules with kinetic modeling to describe gene expression and protein regulation,
while largely relying upon biophysical parameters from the literature.
This formulation significantly reduced the size and complexity of the model compared to the previous study of Tasseff et al., while increasing the breadth of the biology described \cite{Tasseff2011}.
The effective model, despite its simplicity, captured key features of ATRA induced differentiation of HL-60 cells.
Model analysis predicted the bistability of MAPK activation as a function of ATRA exposure; conformational experiments supported ATRA-induced bistability.
Model simulations were also consistent with measurements of the influence of MAPK inhibitors, and the failure of BLR1 knockout cells to differentiate when exposed to ATRA.
Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex.
Taken together, these findings when combined with other literature evidence,
suggested that positive feedback architectures are central to differentiation programs generally, and necessary for ATRA-induced differentiation.

>>>>

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained MEK-dependent activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
and the formation of an ATRA-inducible signalsome complex at the membrane which drives MAPK activation through a yet to be identified kinase activity.
While the makeup of the signalsome complex is not yet known,
we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
Signalsome formation and activity is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1, identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}, and is also involved with signalsome activity.
Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
A previous computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
However, this previous study developed and analyzed a complex model,
thus leaving open the critical question of what is the minimal positive feedback circuit required to drive ATRA-induced differentiation.

In this study, we explored this question using a minimal mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
The ATRA responsive signalsome-MAPK circuit was then used to drive a downstream gene expression program which encoded for the expression of intermediate and functional differentiation markers.
The effective model used a novel framework which integrated logical rules with kinetic modeling to describe gene expression and protein regulation,
while largely relying upon biophysical parameters from the literature.
This formulation significantly reduced the size and complexity of the model compared to the previous study of Tasseff et al., while increasing the breadth of the biology described \cite{Tasseff2011}.
The effective model, despite its simplicity, captured key features of ATRA induced differentiation of HL-60 cells.
Model analysis predicted the bistability of MAPK activation as a function of ATRA exposure; conformational experiments supported ATRA-induced bistability.
Model simulations were also consistent with measurements of the influence of MAPK inhibitors, and the failure of BLR1 knockout cells to differentiate when exposed to ATRA.
In addition, the expression of intermediate and phenotypic differentiation markers as also captured following ATRA exposure.
Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex.
Taken together, these findings when combined with other literature evidence,
suggested that positive feedback architectures are central to differentiation programs generally, and necessary for ATRA-induced differentiation.
